Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-07-08 motivation.tex

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, reading the same value for
  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
  with same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
when executed with a \emph{Read Committed} (RC) isolation level, the
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e,, both transactions modifying
the same record) among concurrent transactions when they
commit. Write-write conflicts can be eliminated in various ways, e.g.,
through conflict detection followed by a rollback, or through
exclusive locks, or a combination of both.  For instance, one possible
implementation of SI, close to the one used by
PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

>>>>

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e,, such a conflict occurs if
concurrentlyh executing transactions modify the same record) among
concurrent transactions when they commit. Write-write conflicts can be
eliminated in various ways, e.g., through conflict detection followed
by a rollback, or through exclusive locks, or a combination of both.
For instance, one possible implementation of SI, close to the one used
by PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

### 2017-07-08 opsem.tex

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction comes
tagged with an isolation specification $\I$, which has two components
$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
that the transaction can witness, during its execution ($\I_e$), and
when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
predicates over the (current) transaction-local database state
($\stl$), the state ($\stg$) of the global database when the
transaction last took a step, and the current state ($\stg'$) of the
global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in (Sec.~\ref{sec:motivation}), an SI transaction on
PostgreSQL defines $\I$ as follows:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

@@

\textbf{Local Context Independence} As mentioned previously, our
operational semantics does not let a transaction read its own writes.
It also does not let a transaction overwrite its own writes, due to
the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on the
\rulelabel{E-Delete} and \rulelabel{E-Update} rules. We refer to this
restriction as \emph{local context independence}.  This restriction is
easy to relax in the operational semantics and the reasoning framework
presented in the next section; the automation framework introduced in
Sec.~\ref{sec:inference}, however, has a non-trivial dependence on
this assumption. Nonetheless, we have encountered few instances in
practice where enforcing local context independence turns out to be a
severe restriction. Indeed, all 17 transactions in the TPC-C and TPC-E
benchmarks satisfy this assumption.

>>>>

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction comes
tagged with an isolation specification $\I$, which has two components
$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
that the transaction can witness, during its execution ($\I_e$), and
when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
predicates over the (current) transaction-local database state
($\stl$), the state ($\stg$) of the global database when the
transaction last took a step, and the current state ($\stg'$) of the
global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in \S\ref{sec:motivation}, an SI transaction on
PostgreSQL defines $\I$ as follows:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

@@

\textbf{Local Context Independence} As mentioned previously, our
operational semantics does not let a transaction read its own writes.
It also does not let a transaction overwrite its own writes, due to
the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on the
\rulelabel{E-Delete} and \rulelabel{E-Update} rules. We refer to this
restriction as \emph{local context independence}.  This restriction is
easy to relax in the operational semantics and the reasoning framework
presented in the next section; our inference procedure described in
\S\ref{sec:inference}, however, has a non-trivial dependence on
this assumption. Nonetheless, we have encountered few instances in
practice where enforcing local context independence turns out to be a
severe restriction. Indeed, all 17 transactions in the TPC-C and TPC-E
benchmarks satisfy this assumption.
